Assignment 1
============

## 1. Purpose

The purpose of the assignment is to analyse the activity of an individual monitored through a device over a two-month period. The following questions will be asnwered:

(1) What is mean total number of steps taken per day?
(2) What is the average daily activity pattern?
(3) Are there differences in activity patterns between weekdays and weekends?

## 2. Analysis

Let's load and take a look at the data set.

```{r echo = TRUE}

activityData <- read.csv("D:/Data science specialisation/8. Reproducible research/activity.csv")
summary(activityData)

```

Note that the steps data has some gaps (indicated by NAs). We will come back to that later. For now, we will ignore those when calculating statistics.

###(1) What is mean total number of steps taken per day?

We now calculate the total number pf steps,and mean and median number of steps per day.

```{r echo = TRUE, results='hide'}
library(dplyr)

timeIntervalOrder <- numeric()

#label each time interval in terms of its order within a single day
#this will be used to asnwer question 2
for (i in 1:nrow(activityData)) {
        
        timeIntervalOrder[i] <- sum(activityData$date[1:i] == activityData$date[i])
        
        
}

activityData <- cbind(activityData,timeIntervalOrder)
        
activityDataClean <- activityData[complete.cases(activityData),]

byDate <- activityDataClean %>% group_by(date) %>% summarise(totalSteps = sum(steps))
                                                             


totalSteps <- sum(byDate$totalSteps)

meanDailySteps <- mean(byDate$totalSteps)

medianDailySteps <- median(byDate$totalSteps)
```

The total number of steps made over the two-month period is `r format(totalSteps, nsmall = 0)`. The average number of steps a day is `r format(meanDailySteps,nsmall = 0)` while the median is `r format(medianDailySteps, nsmall = 0 )`.

Here is how daily steps are distributed.

```{r echo = TRUE}
library(ggvis)

g <- byDate %>% ggvis(~totalSteps) %>% layer_histograms()
g

```

As mentioned before, there are a number of missing values in the data set. In the following seciton, we fill-in those missing values using the average number of steps per interval and see how the statistics we calculated so far will change.

```{r echo = TRUE, results='hide'}

activityDataImputed <- activityData

for (i in 1:nrow(activityData)) {
 
        if (is.na(activityDataImputed$steps[i])) {
                
                activityDataImputed$steps[i] <- mean(activityData$steps, na.rm = TRUE)
        }
               
}

byDateImputed <- activityDataImputed %>% group_by(date) %>% summarise(totalSteps = sum(steps))


totalStepsImputed <- sum(byDateImputed$totalSteps)

meanDailyStepsImputed <- mean(byDateImputed$totalSteps)

medianDailyStepsImputed <- median(byDateImputed$totalSteps)


```

The total number of steps made over the two-month period is `r format(totalStepsImputed, nsmall = 0)` - a `r format(totalStepsImputed/totalSteps*100 - 100, nsmall = 0)`% change compared to previos result when we ignored the missing values. The average number of steps a day is `r format(meanDailyStepsImputed,nsmall = 0)` while the median is `r format(medianDailyStepsImputed, nsmall = 0 )` - very small change compared to the previous values.

Here is how daily steps are distributed.

```{r echo = TRUE}

g3 <- byDateImputed %>% ggvis(~totalSteps) %>% layer_histograms()
g3

```

###(2) What is the average daily activity pattern?

Next, we find out which time of the day was when the greatest activity took place.

```{r echo = TRUE, results='hide'}

byInterval <- activityDataClean %>% group_by(timeIntervalOrder) %>% summarise(meanSteps = mean(steps))

g2 <- byInterval %>% ggvis(~timeIntervalOrder,~meanSteps) %>% layer_lines()

maxStepsInInterval <- max(byInterval$meanSteps)
mostActiveInterval <- grep(maxStepsInInterval,byInterval$meanSteps)
```

The most active time of the day on average is `r format(mostActiveInterval*5/60, nsmall = 2, digits = 3)` o'clock when there were `r format(maxStepsInInterval, nsmall = 0, digits = 3)` steps were made.

The following graph shows the distribution of the number of steps made on average in which 5-minute time interval.

```{r echo = FALSE}
g2

```

###(3) Are there differences in activity patterns between weekdays and weekends?

```{r echo = TRUE,results='hide'}

activityDataImputed$date <- as.Date(activityDataImputed$date)

t <- weekdays(activityDataImputed$date, abbreviate = TRUE)

weekTime <- vector()

for (i in 1:nrow(activityDataImputed)) {

        if (t[i] == "Sat" || t[i] == "Sun") {
        
          weekTime[i] <- "weekend"
        } else {
        
               weekTime[i] <- "weekday"
        
        }
        
}

activityDataImputed <- cbind(activityDataImputed, weekTime)

weekEndDays <- activityDataImputed[,"weekTime"] == "weekend"

byIntervalImputedWeekend <- activityDataImputed[weekEndDays,] %>% group_by(timeIntervalOrder) %>% 
                                                                                summarise(meanSteps = mean(steps))

byIntervalImputedWeekday <- activityDataImputed[!weekEndDays,] %>% group_by(timeIntervalOrder) %>% 
                                                                                summarise(meanSteps = mean(steps))


g4 <- byIntervalImputedWeekend %>% ggvis(~timeIntervalOrder,~meanSteps) %>% layer_lines()
g5 <- byIntervalImputedWeekday %>% ggvis(~timeIntervalOrder,~meanSteps) %>% layer_lines()



g5
```

The following two graphs show the activity levels during weekends and weekdays via plotting average number of steps over 5-minute intervals.

Weekends activity
```{r echo = FALSE}
g4
```

Weekdays activity
```{r echo = FALSE}
g5
```